\bf Model Visualization Techniques for a Social Network Model

### 2018-05-09 paper-child.Rnw
going to submit this week. I'm determined :muscle:

\par In this paper, we concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, where other models only rely on the network statistics, to model the changes in the network. Allowing the actor-level statistics to directly effect the structure of the network leads to a more practical and relevant approach to model change in a social network: we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this intuition in the modeling process.

%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. They are relatively new, especially compared to the classic ERGMs, and they are not very tractable analytically. Likelihood functions quickly become very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit models, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this,
SAOMs are often fit to data using a series of Markov Chain Monte Carlo (MCMC) phases for finding method of moments estimates of model parameters. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user. In this paper, in order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out of their black boxes. %, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By bringing some light to the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.

\par We get into the SAOM black box by using model visualization (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the process of fitting the SAOMs as opposed to looking at only the final output. Stochastic actor-oriented models are a prime example of a set of models that can benefit greatly from the application of model visualization. For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process. Bringing the CTMC out of the black box and into the light through model visualization can provide researchers with insights into the underlying features of the model. Furthermore, SAOMs can include a great deal of parameters to be added to the model structure, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, which causes high correlation between the associated parameters in a SAOM. By visualizing collections of SAOMs, we gain a better understanding of these correlations and find ways to deal with them and rectify their effects in the model. In addition, the estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and the convergence checks for these estimates rely on simulation from the fitted model. Again, each of these steps are largely kept in the background of the estimation process. With the help of static and dynamic visualizations we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.

@@

@@

Network data is of frequent interest to researchers in a wide array of fields. There are technological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, like neural networks, and social networks, just to name a few \citep{newman}. Each of these examples have one thing in common: their data \textit{structure}. There are always units of observation: the power stations, websites, neurons, and people, which we refer to throughout this paper as \textit{nodes} or \textit{actors}. There are also always connections of some kind between those units: the power lines, hyperlinks, electrical signals, and relationships, which we will call \textit{edges} or \textit{ties}. Networks might change over time, like when new websites and hyperlinks are added on the World Wide Web, or when there are new people and relationships in a friendship network. The nodes and edges themselves can also have inherent variables of interest, e.g.\ the institution of authors in a co-authorship network, or the number of times two authors have collaborated.

The multiple layers of network data structures pose unique problems to network analysts. Some questions that network researchers may aim to answer are: How does the strength of a tie between two nodes affect the overall structure of the network? Do node-level differences affect the formation or dissolution of edges? Which views of the data are most informative for communicating significant effects and other results of statistical analyses of the network of interest? These are, of course, just a few broad questions, and we focus here on the latter, which we aim to answer through visual exploration of network data and models.

@@

Network visualization, also called network mapping, is a prominent subfield of network analysis. Visualizing network data is uniquely difficult because of the structure of the data itself.  Most, if not all, data visualizations rely on well-defined axes inherited from the data. If variables are numerical, histograms, scatterplots, or time series plots are straightforward to construct. If the variables are categorical, bar charts and mosaic plots are available to to the researcher. If the data are spatial, there is a well-defined region in which to view information. Network data, however, are much less cut-and-dried.

@@

<<toyex, echo = FALSE>>=
set.seed(81393)
nodes <- data.frame(id = 1:5, group = sample(LETTERS[1:2], size = 5, replace = T), stringsAsFactors = F)
g <- igraph::as_data_frame(igraph::random.graph.game(n = 5, p = .4))
toynet <- list(nodes = nodes, edges = rbind(g, c(5,4)))
@

The first method, the node-link diagram, represents nodes with points in 2D Euclidean space and then represents edges by connecting the points with lines when there is an edge between the two nodes. These lines can also have arrows on them indicating the direction of the edge for directed networks. But because there is typically no natural placement of the points unless they have important spatial locations, a random placement of the points is used, then adjusted via a layout algorithm, of which there are many \citep{drawingalgs}.

<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm. On the right, the adjacency matrix visualization for that same network.", out.width='.49\\textwidth'>>=
toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
ggplot(data = toynetdat) +
  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 11, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, fontsize = 9, arrowgap = .05, directed = T) +
  ThemeNet +
  theme(plot.title = element_text(size = 20)) +
  labs(title = "Node-Link Diagram")
#ggsave("figure/toynetnld.pdf",plot = nld, width = 4, height = 4)

@@

\st{The second primary method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$, describes the edges of a network in matrix form. An entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as
\[
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists } i\to j \\
                                   0 & otherwise
  \end{cases}
\]
Note here that our edge variables are binary: we only consider the presence or absence of an edge. If the network has weighted edges, for example an email network where edge weights represent the number of emails sent from one person to another, the entries in the adjacency matrix are the edge weights instead of zeroes and ones. In an undirected network, $A_{ij} = A_{ji}$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$. Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks only if every edge between two nodes is reciprocated.}
%\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.

>>>>

\par We concentrate on one type of model for dynamic social networks: the stochastic actor-oriented models (SAOMs), introduced by \citet{saompaper}. These models are fundamentally different from other social network models because they allow us to incorporate network \textit{and} actor statistics, where other models only rely on the network statistics, to model the changes in the network. Allowing the actor-level statistics to directly effect the structure of the network leads to a more practical and relevant approach to model change in a social network: we expect people with common interests to be more likely to form relationships, and SAOMs allow us to incorporate this intuition in the modeling process.

%\par In order to estimate the parameters of  SAOMs the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, was developed \cite{RSiena}. This software marks a huge contribution to the field of social network analysis., and when we began working with it we were eager to understand how it works. HH: we are still interested in this.
%\hh{The next sentence is very vague - make it more specific: what aspect of SAOMs is not well understood? I think what you want is something like ``SAOMs are not very tractable analytically with traditional methods, e.g. likelihood functions turn into very complex objects because of the sheer number of parameters involved. Therefore, computationally more tractable solutions are used to fit estimators, in particular, SAOMs are fit using ... }
Unlike other network models, SAOMs are not very well understood. They are relatively new, especially compared to the classic ERGMs, and they are not very tractable analytically. Likelihood functions quickly become very complex objects to analyze due to the dependency structure inherent in the data. Therefore, computationally more tractable solutions are used to fit models, and in particular, % Traditional modeling steps, like determining the likelihood function, are very complex. Because of this,
SAOMs are often fit to data using a series of Markov Chain Monte Carlo (MCMC) phases for finding method of moments estimates of model parameters. In order to estimate the parameters of SAOMs, we use the software SIENA, and its \texttt{R} implementation \texttt{RSiena}, which was developed by \citet{RSiena}. This software marks a huge contribution to the field of social network analysis, but the many moving pieces involved in parameter estimation are largely ``behind the scenes" and hidden from the software user. In order to better understand the model-fitting process, we attempt to bring SAOM fits and the fitting process out into the light. %, by combining the principals of network visualization with those of model visualization as discussed in \citet{modelvis}. %Here, we explore RSiena and SAOMs with innovative network visualizations, both static and dynamic.
By visualizing the underlying methods and structures that are behind the scenes when fitting SAOMs to network data, we aim to help researchers working with the models better understand the implications and analyses of these models.

\par To get behind the scenes of the SAOM fitting process, we use model visualization techniques (see \citet{modelvis}) to display the model in the data space, view collections of models instead of single models, and visually explore the process of fitting the SAOMs as opposed to looking at only the final output. Stochastic actor-oriented models are an excellent example of a family of models that can benefit greatly from the application of model visualization. For instance, the models themselves include a continuous-time Markov chain (CTMC) that is completely hidden from the analyst in the model fitting process. Bringing the CTMC out of the black box and into the light through model visualization can provide researchers with insights into the underlying features of the model. Furthermore, SAOMs can include a great deal of parameters to be added to the model structure, each of which is attached to a network statistic. These statistics are often somewhat, if not highly, correlated, which causes high correlation between the associated parameters in a SAOM. By visualizing collections of SAOMs, we gain a better understanding of these correlations and can find ways to account for their effects in the model. In addition, the estimation of the parameters in a SAOM relies on a Robbins-Monro algorithm, and the convergence checks for these estimates rely on simulation from the fitted model. Again, each of these steps are largely kept in the background of the estimation process. With the help of static and dynamic visualizations we bring the hidden model fitting processes into the foreground, eventually leading to a better understanding and higher accessibility of stochastic actor-oriented models for social network analysts.

@@

We provide a brief introduction to the structure of network data and common network visualization methods.

@@

Network data is of frequent interest to researchers in a wide array of fields, such as biologists studying neural networks and sociologists studying social networks.% There are technological networks, like power grids or the internet, information networks, such as citation networks or the World Wide Web, biological networks, like neural networks, and social networks, just to name a few \citep{newman}. 
% Each of these examples have one thing in common:
Network data across fields have a similar data structure. There are always units of observation, such as neurons %the power stations, websites, neurons, 
and people, which we refer to throughout this paper as \textit{nodes} or \textit{actors}. There are also always connections of some kind between those units, such as electrical signals and relationships, which we will call \textit{edges} or \textit{ties}. Networks often change over time, such as when there are relationships formed in a friendship network. The nodes and edges themselves can also have inherent variables of interest, e.g.\ the age, gender, and political affiliation of people in a friendship network and the length of a friendship.  %the institution of authors in a co-authorship network, or the number of times two authors have collaborated.

The multiple layers of network data structures pose unique problems to analysts. Some questions that researchers may aim to answer are: How does the strength of a tie between two nodes affect the overall structure of the network? Do differences in node variables affect the formation or dissolution of edges? Which views of the data are most informative for communicating significant model effects? What about other results of statistical analyses of the network? %These are, of course, just a few broad questions, and we focus here on the latter, which 
We aim to answer some of these questions through visual exploration of network data and models.

@@

Network visualization, also called network mapping, is a prominent subfield of network analysis. Visualizing network data is uniquely difficult because of the structure of the data itself.  Most data visualizations rely on well-defined axes inherited from the data, such as Cartesian coordinates or spatial locations. %If variables are numerical, histograms, scatterplots, or time series plots are straightforward to construct. If the variables are categorical, bar charts and mosaic plots are available to to the researcher. If the data are spatial, there is a well-defined region in which to view information. 
Network data, however, are much less cut-and-dried.

<<toyex, echo = FALSE>>=
set.seed(81393)
nodes <- data.frame(id = 1:5, group = sample(LETTERS[1:2], size = 5, replace = T), stringsAsFactors = F)
g <- igraph::as_data_frame(igraph::random.graph.game(n = 5, p = .4))
toynet <- list(nodes = nodes, edges = rbind(g, c(5,4)))
@
<<toyplots, fig.show='hold', fig.cap="On the left, a node-link diagram of our directed toy network, with nodes placed using the Kamada-Kawai algorithm. On the right, the adjacency matrix visualization for that same network.", out.width='.49\\textwidth', fig.height=4>>=
toynetdat <- fortify(as.edgedf(toynet$edges), toynet$nodes)
ggplot(data = toynetdat) +
  geom_net(aes(from_id=from_id, to_id=to_id), labelon = T, size = 11, labelcolour = 'white', ecolour = 'black', vjust = .5, hjust = .5, fontsize = 9, arrowgap = .05, directed = T) +
  ThemeNet +
  theme(plot.title = element_text(size = 20)) +
  labs(title = "Node-Link Diagram")
#ggsave("figure/toynetnld.pdf",plot = nld, width = 4, height = 4)

@@

The first method, the node-link diagram, represents nodes with points in two dimensions and then represents edges by connecting the points with lines when there is an edge between the two nodes. These lines can also have arrows on them indicating the direction of the edge for directed networks. But because there is typically no natural placement of the points unless they have inherent spatial locations, a random placement of the points is used, then adjusted via a layout algorithm, of which there are many \citep{drawingalgs}.

@@

The second primary method for network visualization uses the adjacency matrix of the network. The adjaceny matrix of a network, $\mathbf{A}$ is defined as follows: an entry $A_{ij}$ of $\mathbf{A}$, for two nodes $i\neq j$ in the network is defined as
\[
  A_{ij} =
  \begin{cases}
                                   1 &  \text{if an edge exists } i\to j \\
                                   0 & otherwise
  \end{cases}
\]
We only consider the presence or absence of an edge, but if the network has weighted edges, for example an email network where edge weights represent the number of emails sent from one person to another, the entries in the adjacency matrix can be the edge weights instead of zeroes and ones. In an undirected network, $A_{ij} = A_{ji}$ for all $i \neq j$, but in a directed graph this is only true if there is an edge from $i$ to $j$ and from $j$ to $i$. Thus, $\mathbf{A}$ is always symmetric for undirected networks, and is symmetric for directed networks if and only if every edge between two nodes is reciprocated.
%\hh{XXX what is an adjacency matrix? - explain at this point in mathematical terms} \st{XXX see above XXX}
An adjacency matrix visualization for our toy example is also shown in Figure~\ref{fig:toyplots}.
